# Subclustering

## Importing libraries

```{r, message = FALSE}
library(Seurat)
library(tidyverse)
library(RColorBrewer)

```

```{r, include=FALSE}
mytheme <-  theme(axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      legend.title = element_text(size = 12, face = "bold"),
      legend.text=element_text( size = 10), 
      text=element_text(family="Arial")) 
```

## Importing Data
```{r, message = FALSE}
hpf18_seurat <- readRDS(file = "rds/hpf18_seurat_optimized551.rds" )

```

## Subclustering the TB cells - Using only cells that are assigned to the 'shared' cluster across all the clustering resolutions.

```{r}
TBcells <- subset(hpf18_seurat, subset = TB == TRUE)
TBcells
```

## Dimensional reduction with UMAP and clustering
```{r}
# Running PCA
TBcells <- RunPCA(TBcells, npcs = 50)

# Elbow plot to determine number of PCs for dimension reduction

ep <- ElbowPlot(TBcells, ndims = 50, reduction = "pca") +
  geom_vline(xintercept = 25,  color = "red", size=1.5, linetype="dotted") +
  ylab(label = "Percentage of variance explained") + 
  theme_bw() + theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    text=element_text(family="Arial"))

# UMAP embedding and clustering
ep
TBcells <- RunUMAP(TBcells, dims = 1:25,  min.dist = 0.7, n.neighbors = 30L, verbose = TRUE)
TBcells <- Seurat::FindNeighbors(TBcells, dims = 1:25, k.param = 20, verbose = FALSE) 
TBcells <- Seurat::FindClusters(TBcells, algorithm = 1, resolution = 1, verbose = FALSE)
```

## Identifying marker genes for each cluster 

```{r subclusters_DEG, fig.width = 12, fig.height = 15 }
# Setting identity for cells to seurat clustering results

TBcells<- SetIdent(TBcells, value = TBcells@meta.data$seurat_clusters) 

all_clusters <- FindAllMarkers(TBcells, test.use = "wilcox", only.pos = TRUE) %>%
  group_by(cluster)%>%
  filter(p_val_adj < 0.05)%>%
  arrange(cluster, desc(abs(avg_logFC)),p_val_adj)

#write.csv(all_clusters,file = "Files/2g)MarkerGenes_9Clusters.csv")

# Viewing number of cells in each cluster

table(TBcells@meta.data$seurat_clusters)

# Plot a heatmap to visualize the top 7 marker genes in each cluster. 

top7_wilcox <- all_clusters %>% group_by(cluster) %>% top_n(7, avg_logFC)
DoHeatmap(object = TBcells, features = as.character(top7_wilcox$gene), label = FALSE) +
   theme(legend.title = element_text(size = 25),
         axis.line=element_blank(),
      axis.text.x=element_blank(),
      axis.text.y = element_text(size = 15),
      axis.ticks=element_blank(),
      axis.title.x=element_blank(),
      axis.title.y=element_blank(),
      legend.text=element_text(face="bold", size = 25), 
      text=element_text(family="Arial"))  

#ggsave("2g)AllClusters_markerGeneHeatmap.tiff", hm, dpi = 500, width = 17, height = 17,units = c("cm"))
```

```{r check_DEG}
top7_wilcox
```

## Merging clusters that do not appear distinct

```{r merge_meso}
# Merging the two mesodermal clusters 1 and 4

meso <- subset(x = TBcells, subset = (seurat_clusters == "1" | seurat_clusters == "5"))
meso_ids <- rownames(meso@meta.data)
TBcells@meta.data$seurat_clusters[rownames(TBcells@meta.data) %in% meso_ids] <- 1

```
## Plotting UMAP and assigning cluster labels
```{r subclusters_names}

labelList <- data.frame(clusterID = c(0,1,2,3,4,6,7), 
                        clusterNames = c("Posterior NT", 
                                         "Posterior PSM",  
                                        "Nascent somites",
                                        "Pou5f3+ Posterior NT",
                                        "Maturation zone", 
                                        "Tail somites", 
                                        "NMps"))

TBcells$subclusters_names <- labelList$clusterNames[match(TBcells$seurat_clusters, labelList$clusterID)]
TBcells$subclusters_names <- factor(TBcells$subclusters_names)
TBcells<- SetIdent(TBcells, value = TBcells@meta.data$subclusters_names)

```

## Checking the marker genes for each cluster

```{r check_subclusters_violin, fig.width = 8, fig.height=12}
marker_genes <- c("sox2", "ta", "pou5f3", "olig4", "zic5","tbx6", "tbx16","msgn1","meox1", "mespab", "cyp26a1", "ripply1")
VlnPlot(object = TBcells, features = marker_genes, ncol = 3)
```

```{r UMAP_subclusters, fig.width = 8, fig.height = 8}
# Plotting UMAP dimplot with cluster labels

cols <- c(scales::hue_pal()(7), "#FFFFFF")

DimPlot(TBcells, label = TRUE, label.size = 5, pt.size=0.6)+
  scale_color_manual(values = cols, name = "Cluster Identities")+
  mytheme

#ggsave("2g)TBsubset.tiff", tbsubset_plt, dpi = 500, width = 17, height = 13,units = c("cm"))
#saveRDS(TBcells, file = "rds/2g)TBcells_24.10.19.rds")
```
## Sox2 tbxta coexpression
```{r sox2ta_coexp}

# Plotting co-expression of sox2 and ta in TBcells

FeaturePlot(TBcells, features = c("sox2", "ta"), blend = TRUE, pt.size = 1, blend.threshold = 0.5)
#ggsave("2g)sox2_ta_coexp.tiff", coexp, dpi = 400, width = 12, height = 6,units = c("cm"))

```
## Viewing genes of interest with a dotplot
```{r specify_genes}
# Looking at only cells in the NM decision

TBcells_nmdecision <- subset(TBcells, subset = subclusters_names == "Neuromesodermal Progenitors" |
                              subclusters_names == "Maturation zone" | 
                              subclusters_names == "Pou5f3+ Posterior NT" |
                              subclusters_names == "Posterior NT")
# Specifying genes to visualize
nm_genes <- c("cyp26a1","eve1","fgf8a","hoxa13b","hoxc13a","hoxc13b","sox2","ta","wnt8a")
meso_genes <- c("aldh1a2", "apoc1l","bmp10", "her1", "her7", "msgn1","snai1a","tbx6","tbx16","tbx24","uncx4.1", "meox1", "mespab")
neural_genes <- c("gfap", "sox3", "prtga", "zic2a", "zic5")

```

```{r nm_dotplot}
DotPlot(TBcells, features = nm_genes) + 
    theme(text = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.text.x = element_text(size = 8, angle = 35))
```

```{r meso_dotplot}
DotPlot(TBcells, features = meso_genes) + 
    theme(text = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.text.x = element_text(size = 8, angle = 35))
```

```{r neural_dotplot}
DotPlot(TBcells, features = neural_genes) + 
    theme(text = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.text.x = element_text(size = 8, angle = 35))
```

*Q1) How would you save the images in the  working directory?* 

## Exporting the violin plots for genes of interest.

We loop over each violin plot using the `lapply` function, which merits additional comment.The apply family of functions is very useful to perform an action repeatedly on a collection of objects such as a dataframe, vector, list etc. Normally iteration is done with a for loop, which is rather time-consuming. The apply function performs the iteration without the explicit use of a for loop. `lapply` returns a list. 

*Q2) What is the difference between lapply and sapply?* 

```{r all_violin}
# Looking at expression of nmp markers over all clusters in individual violinPlots
nmp_markers <- c(
"fgf8a",
"wnt8-2",
"wnt8a",
"ta",
"apela",
"eve1",
"ndnf",
"cyp26a1",
"hoxc13b",
"id3",
"hoxc13a",
"thbs2",
"sp5l",
"tagln3b",
"mcamb",
"her12",
"hoxa13b",
"hoxd12a",
"kif26ab",
"sox2",
"prickle1b",
"si:dkey-261h17.1",
"cdh6",
"cx43.4",
"zgc:158328",
"fndc3ba",
"mnx1",
"hes6",
"znf703",
"ube2e2",
"angptl2b",
"depdc7",
"efhd2",
"kdm6a",
"nradd",
"enc1",
"snai1a",
"arf1",
"flrt3",
"phgdh",
"traf4a",
"wls",
"sepn1",
"esrrga",
"ptmab",
"ptbp1b",
"ptmaa",
"mgat1b",
"ubl3a",
"sall4",
"tuba8l4",
"hoxd13a",
"LOC100537138",
"nog2",
"h3f3a",
"myclb",
"sult6b1",
"galnt2",
"fgf4",
"foxa",
"bbc3",
"hoxd11a",
"rcn3",
"plk4",
"foxd3",
"pip5k1ca",
"sec22bb",
"slc12a7b",
"marveld1",
"pcp4a"
)

#write.table(nmp_markers, file = "nmp_markergenes.txt")
plots <- VlnPlot(object = TBcells, features =nmp_markers, combine = FALSE)
plots <- lapply(X = plots, FUN = function(x) x +  theme(plot.title = element_text(size = 30),
                        text = element_text(size = 15),
                        axis.text.y = element_text(size = 12),
                        axis.text.x = element_text(size = 12, angle = 35)))

for (i in 1:length(plots)) {
  ggsave(plot = plots[[i]], path = "violin_plots", filename = paste0("violin_",nmp_markers[i],".png"), width = 14, height = 10, units = "cm")
}

```

## Saving RDS object
```{r}
saveRDS(TBcells, file = "rds/TBcells_subclustered.rds")
```

## Session Info

<details><summary>View Session Info</summary>
```{r, collapse = TRUE}
devtools::session_info()
  
```
</details>

## Solutions

1. Use the ggsave function. 
- ggsave("2g)nm_dotplot.tiff", nm_dotplot, dpi = 400, width = 12, height = 8,units = c("cm"))
- ggsave("2g)meso_dotplot.tiff", meso_dotplot, dpi = 400, width = 12, height = 8,units = c("cm"))
- ggsave("2g)neural_dotplot.tiff", neural_dotplot, dpi = 400, width = 12, height = 8,units = c("cm"))
2. lapply returns a list of the same length as the input vector. sapply (simple apply) returns a vector.

